Towards a standardized evaluation of imputation methodology

### 2022-03-30 Formatted/Formatted.tex
Start writing out the missingness mechanisms part

The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. A popular variant is multiple imputation \citep{rubi76}, whereby each missing value is imputed multiple times. The completed data that are thus obtained can be analyzed by standard techniques and, in the case of multiple imputation, the analysis results can be combined into a single inference \citep[using Rubin's rules][]{rubi87}. In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion, mean imputation, regression imputation, last observation carried forward, indicator method), multiple imputation properly takes into account the sources of uncertainty that are related to the missingness problem.

@@

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values \citep[generally categorized into MCAR, MAR and MNAR mechanisms cf][see Table \ref{table:mech}]{rubi76}.  Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

@@

Another aspect of missingness patterns that may misguide evaluations is the complexity of the patterns across variables. It is hardly reasonable to imagine empirical data with only one incomplete variable, yet some simulation studies rely on univariate missingness patterns anyways. Extrapolating simulation results from a specific missing data pattern to more intricate empirical missingness patterns could lead to suboptimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may conclude that complete case analysis outperforms their imputation method(s). Such a conclusion would, unfortunately, only translate to data that adhere to the same special case \citep[described by][]{fimd}, while biasing inferences in other cases. Another example of a missingness pattern that may inadverted impose assumptions on the evaluations is a monotone pattern \citep[i.e. a pattern with uniformly increasing missingness proportions along variables][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration in all other situations. A final example is multivariate missingness that is generated using stepwise univariate missingness induction. The resulting incomplete data may not have the intended or expected statistical properties \citep[e.g. skewness][]{ampute}, which complicates definitive conclusions about imputation method performance in relation to the data generating mechanism.

@@

The missingness mechanism is usually assumed to be random (MAR) or completely random (MCAR).

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms. [Add PMAR]}
\label{table:mech}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

Fit of the imputation model: [PPC, ref Mingyang, \citep{nguy17, zhao22}.]

>>>>

The idea behind imputation is to impute (fill in) missing values, to obtain a valid estimate of what could have been. A popular variant is multiple imputation \citep{rubi76}, whereby each missing value is imputed multiple times. The completed data that are thus obtained can be analyzed by standard techniques and, in the case of multiple imputation, the analysis results can be combined into a single inference \citep[using `Rubin's rules'][]{rubi87}. In contrast to ad hoc methods for dealing with missing values (e.g. list-wise deletion, mean imputation, regression imputation, last observation carried forward, indicator method), multiple imputation properly takes into account the sources of uncertainty that are related to the missingness problem.

@@

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values \citep[generally categorized into MCAR, MAR and MNAR mechanisms c.f.][see Table \ref{table:mech}]{rubi76}.  Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

@@

Another aspect of missingness patterns that may misguide evaluations is the complexity of the patterns across variables. It is hardly reasonable to imagine empirical data with only one incomplete variable, yet some simulation studies rely on univariate missingness patterns anyways. Extrapolating simulation results from a specific missing data pattern to more intricate empirical missingness patterns could lead to suboptimal advice in practice. For example, if a simulator induces missingness in the outcome variable of their analysis model exclusively, they may conclude that complete case analysis outperforms their imputation method(s). Such a conclusion would, unfortunately, only translate to data that adhere to the same special case \citep[described in][\S 2.7]{fimd}, while biasing inferences in other cases. Another example of a missingness pattern that may inadverted impose assumptions on generalizations is a monotone pattern \citep[i.e. a pattern with uniformly increasing missingness proportions along variables;][]{litt20}. If a simulator investigates iterative imputation methods under monotone missingness, they might believe that the evaluated imputation algorithms converge instantly, while these methods actually require iteration to produce valid results in all other situations. A final example is multivariate missingness that is generated using stepwise univariate missingness induction. The resulting incomplete data may then not have the intended or expected statistical properties \citep[e.g. skewness;][]{ampute}. This complicates any definitive conclusions about imputation method performance in relation to the data generating mechanism.

@@

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms. [Add PMAR?]}
\label{table:mech}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness. \\
\hline
\end{tabular}
\end{center}
\end{table}

The missingness mechanisms that are induced in simulation studies do not necessarily match the true underlying mechanisms in incomplete empirical data, nor the mechanisms that are assumed when imputing such data. Out of the three mechanisms (see Table \ref{table:mech}), MAR missingness is usually investigated in simulations and assumed in practice. Although MCAR might not reasonably be the true missingness generating mechanism and naturally not typically assumed, the mechanism  can be very insightful in the context of imputation methodology evaluation. By contrast, MNAR

% usually assumed to be random (MAR) or completely random (MCAR).

@@

The fit of the imputation model may be verified using posterior predictive checking. [PPC, ref Mingyang, \citep{nguy17, zhao22}.]
